\begin{thebibliography}{999}

    \bibitem{MeasureTheoryCohn}
    Donald L. Cohn.
    \textit{Measure Theory}.
    Birkhäuser Advanced Texts Basler Lehrbücher.
    Birkhäuser, New York, NY, second edition, 2013.

    \bibitem{FollandRealAnalysis}
    Golland, Gerald B.
    \textit{Real analysis:modern techniques and their applications}
    1999

    \bibitem{ProbabilityBillingsley}
    P.~Billingsley, \emph{Probability and Measure}, 3rd~ed., New York: John Wiley \& Sons, 1995.

    \bibitem{KrappWirth2021}
    L.~S. Krapp and L.~Wirth,
    \newblock ``Measurability in the Fundamental Theorem of Statistical Learning,''
    \newblock {\em arXiv preprint arXiv:2102.04775}, 2021.

    \bibitem{UnderstandinMachineLearning}
    S.~Shalev\hyp{}Shwartz and S.~Ben\hyp{}David,
    \emph{Understanding Machine Learning: From Theory to Algorithms}.
    Cambridge University Press, 2014.
    See \S3.1 (PAC Learning), discussion immediately after Definition~3.1
    (“The definition of Probably Approximately Correct learnability contains two approximation parameters…”).

    \bibitem{StatisticalLearningTheory}
    U.~von~Luxburg and B.~Sch{\"o}lkopf,
    \emph{Statistical Learning Theory: Models, Concepts, and Results},
    Manuscript, Max Planck Institute for Biological Cybernetics, T{\"u}bingen, 2008.
    Available at: \url{https://www.tml.cs.uni-tuebingen.de/team/luxburg/publications/StatisticalLearningTheory.pdf}

    \bibitem{LearnabilityDeterministic}
    A.~Blumer, A.~Ehrenfeucht, D.~Haussler, and M.~K.~Warmuth,
    ``Learnability and the Vapnik--Chervonenkis Dimension,''
    \emph{J. Assoc. Comput. Mach.}, vol.~36, pp.~929--965, 1989.
    \url{https://doi.org/10.1145/76359.76371}.


%    \bibitem{MeasureTheoryCohn}
%    D. L. Cohn, \textit{Measure Theory}, Birkh{\"a}user, Boston, 1980.
%
%    %Published paper
%    \bibitem{Akaike}
%    Akaike, H. (1974).
%    A New Look at the Statistical Model Identification.
%    \emph{IEEE Trans. Automat. Contr.}, \textbf{19}, 716--723.
%
%    %Published paper
%    \bibitem{GreengardRokhlin1987}
%    Greengard, L. and Rokhlin V. (1987).
%    A fast algorithm for particle simulations.
%    \emph{J. Comput. Phys.}, \textbf{73}(2), 325--348.
%
%    %book
%    \bibitem{Feller71} Feller, W. (1971)
%    \emph{An Introduction to Probability Theory and Its Applications},
%    Vol.~2, 2nd ed.
%    Wiley, New York.
%
%    %ArXiv preprint
%    \bibitem{Perelman} Perelaman, G. (2002) The entropy formula for the Ricci flow and its geometric applications.
%    ArXiv: \href{https://arxiv.org/abs/math/0211159}{2110.05123}.

    \bibitem{MartinAnthony}
    Martin Anthony and Peter L. Bartlett.
    \textit{Neural Network Learning: Theoretical Foundations}.
    Cambridge University Press, Cambridge, 1999.
    Available: \url{https://doi.org/10.1017/CBO9780511624216}

    \bibitem{MeasureTheoryLeGall}
    J.-F. Le Gall,
    \textit{Measure Theory, Probability, and Stochastic Processes},
    Graduate Texts in Mathematics, vol.~295,
    Springer, Cham, Switzerland, 2022.
    Available: \url{https://link.springer.com/book/10.1007/978-3-031-14205-5}

    \bibitem{HerrlichAxiomOfChoice}
    Herrlich, Horst.
    \textit{Axiom of choice}.
    Lecture Notes in Mathematics, 1876.
    Springer, Berlin, 2006.



    \bibitem{MunkresTopology}
    J. R. Munkres, \textit{Topology}, 2nd ed., Prentice Hall, 2000.


    \bibitem{AnthonyBartlett1999}
    M.~Anthony and P.~L. Bartlett, \emph{Neural Network Learning: Theoretical
    Foundations}, Cambridge University Press, 1999.

    \bibitem{ChernikovSimon2013}
    A.~Chernikov and P.~Simon, \emph{Externally definable sets and dependent
    pairs}, Israel J. Math. \textbf{194} (2013), 409--425.

    \bibitem{DevroyeGyorfiLugosi1996}
    L.~Devroye, L.~Györfi, and G.~Lugosi, \emph{A Probabilistic Theory of Pattern
    Recognition}, Springer, 1996.

    \bibitem{Dolich2004}
    A.~Dolich, \emph{Forking and independence in o-minimal theories}, J. Symbolic
    Logic \textbf{69} (2004), no.~1, 215--240.

    \bibitem{DriesMiller1994}
    L.~van den Dries and C.~Miller, \emph{On the real exponential field with
    restricted analytic functions}, Israel J. Math. \textbf{85} (1994), 19--56.

    \bibitem{Marker2002}
    D.~Marker, \emph{Model Theory: An Introduction}, Graduate Texts in Mathematics,
    vol. 217, Springer, 2002.

    \bibitem{Sauer1972}
    N.~Sauer, \emph{On the density of families of sets}, J. Combin. Theory Ser. A
    \textbf{13} (1972), 145--147.

    \bibitem{Shelah1972}
    S.~Shelah, \emph{A combinatorial problem; stability and order for models and
    theories in infinitary languages}, Pacific J. Math. \textbf{41} (1972),
    247--261.

    \bibitem{TentZiegler2012}
    K.~Tent and M.~Ziegler, \emph{A Course in Model Theory}, Lecture Notes in
    Logic, vol.~40, Cambridge University Press, 2012.

    \bibitem{vandenDries1998}
    L.~van den Dries, \emph{Tame Topology and O-minimal Structures}, London
    Mathematical Society Lecture Note Series, vol. 248, Cambridge University
    Press, 1998.



\end{thebibliography}